Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-22 paper/paper.tex
no predict in the paper

An Anglican program, or \textit{query}, is compiled into a Clojure
function. When inference is performed with a provided algorithm, this
produces a sequence of output values, or \textit{predicts}. Anglican
shares a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican can
look like the following code:
\begin{lstlisting}[style=default]
(defquery model data
  "chooses a distribution
   which describes the data"
  (let [;; Guessing a distribution
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

@@

    ;; Predicting a, b and the distribution
    (predict :a a)
    (predict :b b)
    (predict :d d)))
\end{lstlisting}
This query defines a probability distribution on three predicted
variables, $\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
positive parameters for the type. Samples from this distribution
can be obtained by running the query under one of Anglican's
inference algorithms.

@@

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values
%Clojure objects
suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm}
%and \texttt{fm}
(Anglican counterparts of Clojure \texttt{defn}).
%and \texttt{fn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe} (as well \texttt{predict}
for the output), and can be seamlessly called from inside
Anglican queries, just like functions locally defined within the
same query.

@@

The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithm. \texttt{\$state}
is a Clojure hash map:
\begin{lstlisting}[style=default]
(def initial-state
  "initial program state"
  {:log-weight 0.0
   :predicts []
   :result nil
   ::mem {}
   ::store nil
   ... })
\end{lstlisting}
The full list of map entries depends on the inference
algorithm. Except for transformation of
\texttt{mem}, the memoization form, CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

The most general form of CPS transformation receives an
expression and a continuation, and returns the expression
in CPS form with the continuation potentially called in multiple
tail positions. General expressions can be somewhat voluntarily
divided into several groups:
\begin{itemize}
    \item binding forms --- \texttt{let} and
        \texttt{loop/recur};
    \item flow control --- \texttt{if}, \texttt{when},
        \texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
    \item function applications and \texttt{apply};
    \item probabilistic forms --- \texttt{predict},
        \texttt{observe}, \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
\end{itemize}
Functions that transform general expressions accept the
expression and the continuation as parameters, and are
consistently named \texttt{cps-of-}\textit{form}, for example,
\texttt{cps-of-do}, \texttt{cps-of-store}.

@@

In addition to checkpoints, there are a few other special forms
--- \texttt{predict}, \texttt{store}, \texttt{retrieve},
\texttt{mem} --- which modify program state. These forms are
translated into expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.

\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person]
                       (sample
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{lstlisting}

@@

The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
an appropriate overloading or implementation of the function --- conventionally,
the identifier is a keyword related to the
algorithm name (\texttt{:lmh} for Lightweight
Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
etc.). The second parameter is a query as defined by
the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
the following Clojure code invokes \texttt{infer} on an Anglican query
defined anonymously via the \texttt{query} form:
%If needed, the query
%can be defined anonymously, right in the argument list of a call
%to \texttt{infer}:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query (predict x)) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, while the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. A query may
accept multiple parameters using Clojure's structured binding.
For instance, it may take multiple parameters as components
of an input vector. In this case, the initial value is given
as a structured value, such as a vector, and the components
of this value are matched to the corresponding
parameters of the query via the destructuring mechanism of Clojure.
%The value is destructured via pattern match in case of a structured binding.
For example,
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (predict (sample (normal mean sd))))

@@

Checkpoints can be uniquely named at compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially, however it impacts the
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address, hence the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))
      x (sample dist)]
  (predict x))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces, however the random values should not be
re-used between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

@@

(defquery baz
   (predict (foo)))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(defm same-customer 
  "observe the same customer twice"
  [time-to-arrive-prior
   lunch-delay
   dinner-delay]
  (let [time-to-arrive
          (sample time-to-arrive-prior)]
    (observe (normal time-to-arrive time-sd) 
             lunch-delay)
    (observe (normal time-to-arrive time-sd)
             dinner-delay)
    (predict :time-to-arrive time-to-arrive)))

(defm different-customers
  "observe different customers"
  [time-to-arrive-prior
   lunch-delay
   dinner-delay]
  (let [time-to-arrive-1
          (sample time-to-arrive-prior)
        time-to-arrive-2
          (sample time-to-arrive-prior)]
    (observe (normal time-to-arrive-1 time-sd) 
             lunch-delay)
    (observe (normal time-to-arrive-2 time-sd)
             dinner-delay)
    (predict :times-to-arrive
      [time-to-arrive-1 time-to-arrive-2])))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(defquery deli [time-to-arrive-prior
                lunch-delay
                dinner-delay]
 (let [is-same-customer (sample (flip p-same))
       observe-customer (if is-same-customer
                           same-customer
                           different-customer)]
    (predict :is-same-customer is-same-customer)
    (observe-customer time-to-arrive-prior
                      lunch-delay
                      dinner-delay)))
\end{lstlisting}

Performing inference on query \texttt{deli} predicts the
probability that the same customer visited the deli twice, as
well as probability distributions of average arrival times for
both cases.

@@

\begin{lstlisting}[style=default]
(def N 5000)
(def predicts (map get-predicts
                   (take N (drop N samples))))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(def p-same+ 
     (/ (count (filter :same-customer predicts))
        (double N)))
\end{lstlisting}

@@

In addition to computing the posterior probability that the same
customer visited the deli twice, we may want to know the average
time (or times) to arrive. In Bayesian inference, it is common
to report distributions instead of `most likely' values. We use
predicted values in retrieved samples to approximate the
distributions, and plot distribution histograms for the same
customer visiting twice (Figure~\ref{fig:dist-same}) and
two different customers (Figure~\ref{fig:dist-diff})

\begin{lstlisting}[style=default]
;; single customer                      
(def time-to-arrive+ 
     (map :time-to-arrive
          (filter :same-customer predicts)))
(def mean-to-arrive+ (mean time-to-arrive+))
(def sd-to-arrive+ (std time-to-arrive+))

;; two customers
(def times-to-arrive+ 
     (map :times-to-arrive 
          (filter (complement :same-customer)
                  predicts)))
(def mean-1-to-arrive+
     (mean (map first times-to-arrive+)))
(def sd-1-to-arrive+
     (std (map first times-to-arrive+)))
(def mean-2-to-arrive+
     (mean (map second times-to-arrive+)))
(def sd-2-to-arrive+
     (std (map second times-to-arrive+)))
\end{lstlisting}

>>>>

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
\begin{lstlisting}[style=default]
(defquery model data
  "chooses a distribution
   which describes the data"
  (let [;; Guessing a distribution
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

@@

    ;; Returning the distribution and parameters
    [d a b]))
\end{lstlisting}
This query defines a probability distribution on three
variables, $\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
positive parameters for the type. Samples from this distribution
can be obtained by running the query under one of Anglican's
inference algorithms.

@@

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values
%Clojure objects
suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm}
%and \texttt{fm}
(Anglican counterparts of Clojure \texttt{defn}).
%and \texttt{fn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.

@@

The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithm. \texttt{\$state}
is a Clojure hash map:
\begin{lstlisting}[style=default]
(def initial-state
  "initial program state"
  {:log-weight 0.0
   :result nil
   ::mem {}
   ::store nil
   ... })
\end{lstlisting}
The full list of map entries depends on the inference
algorithm. Except for transformation of
\texttt{mem}, the memoization form, CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

The most general form of CPS transformation receives an
expression and a continuation, and returns the expression
in CPS form with the continuation potentially called in multiple
tail positions. General expressions can be somewhat voluntarily
divided into several groups:
\begin{itemize}
    \item binding forms --- \texttt{let} and
        \texttt{loop/recur};
    \item flow control --- \texttt{if}, \texttt{when},
        \texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
    \item function applications and \texttt{apply};
    \item probabilistic forms --- \texttt{observe},
        \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
\end{itemize}
Functions that transform general expressions accept the
expression and the continuation as parameters, and are
consistently named \texttt{cps-of-}\textit{form}, for example,
\texttt{cps-of-do}, \texttt{cps-of-store}.

@@

In addition to checkpoints, there are a few other special forms
---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
modify program state. These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.

\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person]
                       (sample
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (eye-color 'bill)
    (eye-color 'john)))
\end{lstlisting}

@@

The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
an appropriate overloading or implementation of the function --- conventionally,
the identifier is a keyword related to the
algorithm name (\texttt{:lmh} for Lightweight
Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
etc.). The second parameter is a query as defined by
the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
the following Clojure code invokes \texttt{infer} on an Anglican query
defined anonymously via the \texttt{query} form:
%If needed, the query
%can be defined anonymously, right in the argument list of a call
%to \texttt{infer}:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query x) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, while the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. A query may
accept multiple parameters using Clojure's structured binding.
For instance, it may take multiple parameters as components
of an input vector. In this case, the initial value is given
as a structured value, such as a vector, and the components
of this value are matched to the corresponding
parameters of the query via the destructuring mechanism of Clojure.
%The value is destructured via pattern match in case of a structured binding.
For example,
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (sample (normal mean sd)))

@@

Checkpoints can be uniquely named at compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially, however it impacts the
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address, hence the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))]
  (sample dist))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces, however the random values should not be
re-used between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

@@

(defquery baz
  (foo))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(defm same-customer 
  "observe the same customer twice"
  [time-to-arrive-prior
   lunch-delay
   dinner-delay]
  (let [time-to-arrive
          (sample time-to-arrive-prior)]
    (observe (normal time-to-arrive time-sd) 
             lunch-delay)
    (observe (normal time-to-arrive time-sd)
             dinner-delay)
    [time-to-arrive]))

(defm different-customers
  "observe different customers"
  [time-to-arrive-prior
   lunch-delay
   dinner-delay]
  (let [time-to-arrive-1
          (sample time-to-arrive-prior)
        time-to-arrive-2
          (sample time-to-arrive-prior)]
    (observe (normal time-to-arrive-1 time-sd) 
             lunch-delay)
    (observe (normal time-to-arrive-2 time-sd)
             dinner-delay)
    [time-to-arrive-1 time-to-arrive-2]))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(defquery deli [time-to-arrive-prior
                lunch-delay
                dinner-delay]
  (let [is-same-customer (sample (flip p-same))
        observe-customer (if is-same-customer
                            same-customer
                            different-customer)]
    {:is-same-customer is-same-customer
     :times-to-arrive (observe-customer 
                        time-to-arrive-prior
                        lunch-delay
                        dinner-delay)}))
\end{lstlisting}

Performing inference on query \texttt{deli} computes the
probability that the same customer visited the deli twice, as
well as probability distributions of average arrival times for
both cases.

@@

\begin{lstlisting}[style=default]
(def N 5000)
(def results (map get-result
                   (take N (drop N samples))))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(def p-same+ 
     (/ (count (filter :same-customer results))
        (double N)))
\end{lstlisting}

@@

In addition to computing the posterior probability that the same
customer visited the deli twice, we may want to know the average
time (or times) to arrive. In Bayesian inference, it is common
to report distributions instead of `most likely' values. We use
query results in retrieved samples to approximate the
distributions, and plot distribution histograms for the same
customer visiting twice (Figure~\ref{fig:dist-same}) and
two different customers (Figure~\ref{fig:dist-diff})

\begin{lstlisting}[style=default]
;; single customer                      
(def time-to-arrive+ 
     (map (comp first :times-to-arrive)
          (filter :same-customer results)))
(def mean-to-arrive+ (mean time-to-arrive+))
(def sd-to-arrive+ (std time-to-arrive+))

;; two customers
(def times-to-arrive+ 
     (map :times-to-arrive 
          (filter (complement :same-customer)
                  results)))
(def mean-1-to-arrive+
     (mean (map first times-to-arrive+)))
(def sd-1-to-arrive+
     (std (map first times-to-arrive+)))
(def mean-2-to-arrive+
     (mean (map second times-to-arrive+)))
(def sd-2-to-arrive+
     (std (map second times-to-arrive+)))
\end{lstlisting}
